---
title: "Correspodence_Analysis_and_Bi_Plots"
author: "Mr. Niteeenkumar"
format: html
editor: visual
---

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:\

## Q.1

```{r}

data1=matrix(c(21,241,251,17,54,40,10,74,65,6,11,8,11,79,108),nrow=5,ncol=3,byrow = TRUE)
rownames(data1)=c('Married','Widowed','Divorced','seperated','Never married')
colnames(data1)=c('Dull',"Routine",'Excited')
data1

```

## a) Calculate Pearson Chi-square statistic

The Chi-square test tells us whether there is a significant association between the variables.

The null hypothesis ($H_0$) : There is no association between the variables.

```{r}
#| echo: false

# Chi-square test
result <- chisq.test(data1)
print(result)

cat("Conclusion based on chi square test:",ifelse(result$p.value < 0.05 ,"There is an association between variables","there is no association betweeen variables"),'\n')

r=rowSums(data1)
c=colSums(data1)

N=sum(r)
cat("Grand Total:",N,'\n')


```

The `echo: false` option disables the printing of code (only output is displayed).

## b) & c) CA Calculations \* Total Inertia

```{r}

# Correspondence Matrix (P)
P =data1/N;P

# marginal probabilities
r_mass=rowSums(P);sum(r_mass)
c_mass=colSums(P);sum(c_mass)

# Diagonal matrices of the inverse square root of masses
(Dr_inv=diag(1/sqrt(r_mass)))
(Dc_inv=diag(1/sqrt(c_mass)))

# matrix of expected proportions
(E_prop=outer(r_mass,c_mass))

# Standardised Residual matrix
S=Dr_inv%*%(P-E_prop)%*%Dc_inv;S

# SVD on S
# S =U*D_alpha*V'

svd_S=svd(S)
U=svd_S$u;  V=svd_S$v;  D=diag(svd_S$d) #matrix of singular values (alpha)

# calculate Principal Inertias (Eigenvalues)
# Inertia(lambda) =(singular value)^2

(eigenvalues= (svd_S$d)^2)

#  c) Obtain Inertia
(Inertia=sum(eigenvalues))
cat("Total Inertia lambda:",Inertia,'\n')

# ----- key Identity Check ------
# Total inertia should also equal to (chisq / N)
identical(round((result$statistic/N),4),round(Inertia,4))
(inertia_check=result$statistic/N)
cat("Chi-square inertia check:",inertia_check,'\n')
cat("Inertia obtained is:",Inertia,'\n')

```

***INTERPRETATION:** This value (Inertia : 0.034) represent total amount of association (variance) in the variables. it's measures how much observed variables ( data) deviates from independence.*

## e) Quality of Representation: Bi-plot

```{r}

# percentage of inertia for each dimension
(perc_inert= eigenvalues/Inertia *100)

# cumulative percentage inertia
(cum_perc_inertia=cumsum(perc_inert))

# Note : the number of dimensions = min(R-1,C-1) = min(4,2)= 2

(inertia_summary=data.frame(
  Dimension=1:length(eigenvalues),
  inertia=eigenvalues,
  percentage_inertia=perc_inert,
  cumulative_perc=cum_perc_inertia
))

```

***Interpretation:** The cumulative percentage for Dimension 2 tells us total inertia captured by our 2D biplot*. in this scenario 100% inertia captured by 2 dimensions this means biplot is perfect representation of data.

## d) Visualization of Bi-plot

*we will use symmetric biplot, where both row and column points are scaled to show their relationship.*

```{r}

# Row principal coordinates (phi)
phi = Dr_inv%*%U%*%D

# Column principal coordinates (gamma)
gamma = Dc_inv%*%V%*%D

rownames(phi)=rownames(data1)
colnames(phi)=paste("Dim",1:ncol(phi))
rownames(gamma)=colnames(data1)
colnames(gamma)=paste("Dim",1:ncol(gamma))

# creating bi-plot using base R
x_lim=range(c(phi[,1],gamma[,1]))*1.2
y_lim=range(c(phi[,2],gamma[,2]))*1.2


plot(phi[,1:2],xlim=x_lim,ylim=y_lim,col='blue',pch=15,
     main="symmetric biplot",
     xlab = paste0("Dimension 1 (", round(perc_inert[1], 1), "%)"),
     ylab = paste0("Dimension 2 (", round(perc_inert[2], 1), "%)") )
# Add row labels
text(phi[,1:2],labels=rownames(phi),col='blue',pos=4) # pos=4 put text to right
# Add column points 
points(gamma[,1:2],col='red',pch=18)
# Add column labels
text(gamma[,1:2],labels=rownames(gamma),pos=1,col='red')
# Add origin lines
abline(h=0,v=0,lty=2,col='gray')
```

***Interpretation from Bi-plot:***

-   **Rules for Interpretation:** Points that are closer to each other has stronger than average association. points that are far apart are disassociated.

1\) ***Routine** and **Divorced are extremely close to each other this shows a strong association.***

2\) "**Dull"** & **"Seperated" these two points are grouped together on far right i.e, seperated individuals are much more likely to report Dull.**

3\) **"Married" near to origin this means the married group attitude profile is very special and doesn't deviate strongly toward any single attitude.**

```{r verification }
library(ca)
(ca_result=ca(data1))
summary(ca_result)
plot(ca_result)
```

------------------------------------------------------------------------

------------------------------------------------------------------------

## Q.2 CAvariants - Selikoff's data

```{r}
library("CAvariants")
data(package='CAvariants')
#| Selikoff's data, a two-way contingency table
#| asbestos
data("asbestos");View(asbestos)

```

## a) Testing the significance of association between variables

```{r}
result2=chisq.test(asbestos,simulate.p.value =TRUE,B=1000);print(result2)

cat("With the help of Non-Parametric test of association it is observed that:",ifelse(result2$p.value <= 0.05 ,"There is an association between variables","There is no association between variables"),'\n')
```

***simulate.p.value =T tells R to use simulation when we have large data to use fisher exact test** & **B = 10000 are number of replicates (for precision)***

## b) Correspondence Analysis to Study Pattern of Association

```{r}

N1=sum(asbestos)

P1= asbestos/N1

r1_mass=rowSums(P1)
c1_mass=colSums(P1)

Dr1_inv=diag(1/sqrt(r1_mass))
Dc1_inv=diag(1/sqrt(c1_mass))

# matrix of expected proportion
E1_prop=outer(r1_mass,c1_mass)

# matrix of standardised residual
S1=Dr1_inv%*%(P1 - E1_prop)%*%Dc1_inv

# SVD of matrix of standardised residual

svd_S1=svd(S1)

D1=diag(svd_S1$d)
eigenvalues_1=(svd_S1$d^2)
Inertia_1=sum(eigenvalues_1)
cat("Total Iertia lambda is:",Inertia_1,'\n')

U1=svd_S1$u ; V1=svd_S1$v

# --- Inertia Check -----
cat("Inertia cross checking , Inertia from result2",round((result2$statistic/N1),3))
cat("INertia obtained from CA:",round(Inertia_1,3))

```

```{r}

(data_summary=data.frame(
             Dimension=1:length(eigenvalues_1),
             Inertia=eigenvalues_1,
             perc_inertia=eigenvalues_1/sum(eigenvalues_1)*100,
             cumulative_perc_inertia=cumsum(eigenvalues_1/sum(eigenvalues_1)*100)
))
```

## c) Bi-plot (2-Dimensional display)

```{r}

phi1=Dr1_inv%*%U1%*%D1
gamma1=Dc1_inv%*%V1%*%D1

rownames(phi1)=rownames(asbestos)
colnames(phi1)=paste("Dim",1:nrow(asbestos))
rownames(gamma1)=colnames(asbestos)
colnames(gamma1)=colnames("Dim",1:row(asbestos))

xlim1=range(c(phi1[,1],gamma1[,1]))*1.2
ylim1=range(c(phi1[,2],gamma1[,2]))*1.2

plot(phi1[,1:2],xlim=xlim1,ylim=ylim1,main="Symmetric Bi-plot",
    xlab=paste0("Dimension 1 (", round(data_summary$perc_inert[1], 1), "%)"),
    ylab=paste0("Dimension 2(",round(data_summary$perc_inertia[2],1),"%)"),
    col='blue',pch=18)
text(phi1[,1:2],labels = rownames(phi1),col='blue',pos=4)
points(gamma1[,1:2],col='red',pch=10)
text(gamma1[,1:2],labels=rownames(gamma1),col='red',pos=1)
abline(h=0,v=0,col='gray')
```

## Cross checking Bi-plot

```{r}
library(ca)
ca_result2=ca(asbestos)
summary(ca_result2)
plot(ca_result2)
```
